---
title: "Meadow Plot Tutorial: Rmd Version"
author: "Paige Duffin"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---
# Overview

This tutorial uses windowed nucleotide diversity (π) and FST output from `pixy`
to create meadow plots for pairwise population comparisons. Each plot displays
windowed FST across chromosomes, with points colored by the log2 π ratio between
the two populations.

The workflow:

1. calculates the log2 π ratio for each comparison,
2. combines π ratios with windowed FST estimates,
3. identifies candidate outlier windows, and
4. creates and stacks the meadow plots.

Candidate outliers are identified separately for each comparison as windows in
the upper 1% of FST and either the lower or upper 1% of the log2 π ratio.
Negative FST estimates are removed before these thresholds are calculated.

If π equals zero in either population, the resulting infinite log2 ratio is
retained and displayed as an asterisk.

## Expected files and directory structure

This tutorial includes four example input files that can be used to run the complete workflow:

- `example_pixy_fst_data.csv`: windowed Fst output data containing `pop1`, `pop2`,
  `chromosome`, `window_pos_1`, `window_pos_2`, and `avg_wc_fst`
- `example_pixy_pi_data.csv`: windowed π output data containing `pop`,
  `chromosome`, `window_pos_1`, `window_pos_2`, and `avg_pi`
- `example_comparison_map.txt`: defines the comparison names (`comp`), population pairs 
  (`population_1`, `population_2`), and plotting order (`plot_order`)
- `example_chromosome_map.txt`: connects chromosome names in the pixy output (`chromosome`) to 
  the numeric labels used in the plots (`chromo.num`)

The example project uses the following structure:

meadow_plot_tutorial.Rmd
input/
  example_pixy_fst_data.csv
  example_pixy_pi_data.csv
  example_comparison_map.txt
  example_chromosome_map.txt

Users should first run the tutorial with the example files and then replace them with files 
for their own dataset. The replacement files may have different names, but they must retain 
the expected column structure described below. Their paths are specified in the `file-paths` 
chunk. 

## Required user modifications

Two code chunks must be updated for a new dataset:

(required) `file-paths`: replace the example paths with the full paths to your input
files and desired output folders.
(optional) `example-plot`: uncomment the provided line of code and replace "Sal.v.SPW"
with a comparison name from the comp column of your comparison map.

Additional aesthetic choices (e.g., plotting colors, point sizes, text sizes, and 
output dimensions) can also be modified throughout the document, but these changes 
are optional.

# 1. Setup

The packages below provide the data-wrangling, plotting, and figure-assembly
tools used throughout the tutorial. Install any missing packages once with
`install.packages("package_name")`, then load them with the below chunk.

This setup chunk also defines the default code-chunk options and a transparent
plot theme used later in the tutorial. Functions from `knitr`, `scales`, `grid`,
and `gtable` are called explicitly with `package::function()` notation.

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(tibble)
library(gridExtra)
library(ggtext)

knitr::opts_chunk$set(echo = TRUE,message = FALSE,warning = FALSE,fig.align = "center")

transparent_theme <- theme(
  panel.grid = element_blank(),
  panel.background = element_blank(),
  plot.background = element_rect(fill = "transparent", color = NA),
  legend.position = "none")
```

## Set file paths

Enter the full path to each input file and choose where the processed data and
plots should be saved. Using full paths allows the tutorial to run regardless
of the current working directory.

The output folders will be created automatically. The final checks stop the 
analysis early if any required input file cannot be found.

Tip: In the line below, R looks for `example_pixy_fst_data.csv` inside an 
`input` subdirectory within the project directory specified by `project_dir`:
`fst_file <- file.path(project_dir, "input", "example_pixy_fst_data.csv")`
Make sure your directory/file structure matches this expectation.

```{r file-paths}
project_dir <- "/Users/paigeduffin/Documents/other/build_meadow.plot_tutorial" # edit to match your absolute file path to the project directory

fst_file            <- file.path(project_dir, "input", "example_pixy_fst_data.csv") # edit to match your file name
pi_file             <- file.path(project_dir, "input", "example_pixy_pi_data.csv") # edit to match your file name
comparison_map_file <- file.path(project_dir, "input", "example_comparison_map.txt") # edit to match your file name
chromo_map_file     <- file.path(project_dir, "input", "example_chromosome_map.txt") # edit to match your file name

wrangled_dir <- file.path(project_dir, "wrangled_data")
plot_dir     <- file.path(project_dir, "plots")

dir.create(wrangled_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(plot_dir, showWarnings = FALSE, recursive = TRUE)

if (!file.exists(fst_file)) stop("Fst input not found: ", fst_file)
if (!file.exists(pi_file)) stop("Pi input not found: ", pi_file)
if (!file.exists(comparison_map_file)) stop("Pairwise comparisons map input not found: ", comparison_map_file)
if (!file.exists(chromo_map_file)) stop("Chromosome map input not found: ", chromo_map_file)
```

## Chromosomes & population comparison files

The comparison and chromosome maps tell the tutorial which comparisons and
chromosomes to include and how to arrange them in the final figure.

Each row of the comparison map defines one pairwise comparison using four
columns: `comp`, `population_1`, `population_2`, and `plot_order`. Population
names must exactly match those in the pixy files. `population_1` is the
numerator and `population_2` is the denominator when calculating the π ratio.
Comparison names should follow the format `population_1.v.population_2` so they
can be separated correctly for the plot legends.

The chromosome map must contain `chromosome` and `chromo.num`. Values in
`chromosome` must match the chromosome names in the pixy files, while `chromo.num`
provides the numeric chromosome labels and plotting order.

```{r analysis-settings}
comparison_map <- read_tsv(comparison_map_file, show_col_types = FALSE)
chromosome_map  <- read_tsv(chromo_map_file, show_col_types = FALSE)

# Extract values used by the remaining analysis
chromosome_accessions <- chromosome_map %>% arrange(chromo.num) %>% pull(chromosome)

comps <- comparison_map$comp

# Order used in the final stacked figures.
plot_order <- comparison_map %>% arrange(plot_order) %>% pull(comp)
```

# 2. Read the pixy output

This section imports the windowed Fst and π files and checks that they contain
the columns required by the remaining analysis. If a required column is
missing, the tutorial stops and reports its name.

The example files are comma-separated and are read with `read_csv()`. If your
pixy output is tab-separated, replace `read_csv()` with `read_tsv()` for both
files.

```{r read-data}
fst_df <- read_csv(fst_file, show_col_types = FALSE)
pi_df  <- read_csv(pi_file, show_col_types = FALSE)

required_fst_columns <- c("pop1", "pop2", "chromosome", "window_pos_1", "window_pos_2", "avg_wc_fst")
required_pi_columns <- c("pop", "chromosome", "window_pos_1", "window_pos_2", "avg_pi")

if (!all(required_fst_columns %in% names(fst_df))) {
  stop("The Fst file is missing: ",paste(setdiff(required_fst_columns, names(fst_df)), collapse = ", "))}

if (!all(required_pi_columns %in% names(pi_df))) {
  stop("The pi file is missing: ",paste(setdiff(required_pi_columns, names(pi_df)), collapse = ", "))}
```

# 3. Prepare the Fst data

This section matches the Fst data to the chromosome and comparison maps. A
standardized population-pair key is created by alphabetizing each pair, allowing
comparisons to be matched regardless of whether a population appears in the
`pop1` or `pop2` column of the pixy output.

Chromosomes not included in the chromosome map are removed, and `chromo.num`
is added for chromosome labeling and plot order. The final check confirms that
Fst data were found for every requested comparison.

```{r prepare-fst}
pair_key <- function(pop_a, pop_b) {paste(pmin(pop_a, pop_b), pmax(pop_a, pop_b), sep = "__")}

comparison_map <- comparison_map %>% mutate(pair_key = pair_key(population_1, population_2))

fst_df_prepared <- fst_df %>%
  mutate(chromo.num = match(chromosome, chromosome_accessions),pair_key = pair_key(pop1, pop2)) %>%
  filter(!is.na(chromo.num)) %>%
  inner_join(comparison_map %>% select(comp, pair_key),by = "pair_key") %>%
  select(chromosome, chromo.num, window_pos_1, window_pos_2, comp,avg_wc_fst)

missing_fst_comparisons <- setdiff(comps, unique(fst_df_prepared$comp))
if (length(missing_fst_comparisons) > 0) {
  stop("No Fst rows were found for: ", paste(missing_fst_comparisons, collapse = ", "))}

# check to make sure this df looks good so far
# see knitted document version for tutorial example output here
glimpse(fst_df_prepared)
```

# 4. Calculate π ratios

The π data are first converted from long to wide format so that each population
has its own column. The code then checks that every population listed in the
comparison map is present in the π dataset.

For each comparison, the π ratio is calculated as `population_1 / population_2`
and then log2-transformed. The chromosome map is joined to the resulting table,
and chromosomes without a matching `chromo.num` value are removed.

```{r prepare-pi}
pi_df_wide <- pi_df %>%
  select(pop, chromosome, window_pos_1, window_pos_2, avg_pi) %>%
  pivot_wider(names_from = pop, values_from = avg_pi)

expected_populations <- unique(c(comparison_map$population_1,comparison_map$population_2))

missing_pi_populations <- setdiff(expected_populations,names(pi_df_wide))

if (length(missing_pi_populations) > 0) {
  stop("The pi file contains no data for: ",
    paste(missing_pi_populations, collapse = ", "))}

# Calculate the pi ratio for every comparison in comparison_map
pi_df_ratios <- lapply(seq_len(nrow(comparison_map)), function(i) {
  
  comparison_name <- comparison_map$comp[[i]]
  population_1 <- comparison_map$population_1[[i]]
  population_2 <- comparison_map$population_2[[i]]
  
  pi_df_wide %>%
    transmute(chromosome,window_pos_1,window_pos_2,
      comp = comparison_name,
      population_1 = population_1,population_2 = population_2,
      pi_ratio = .data[[population_1]] / .data[[population_2]]
    )
}) %>%
  bind_rows() %>%
  left_join(chromosome_map, by = "chromosome") %>%
  mutate(log2_pi.ratio = log2(pi_ratio)) %>%
  filter(!is.na(chromo.num))

# check to make sure this df looks good so far 
# see knitted document version for tutorial example output here
glimpse(pi_df_ratios)
```

# 5. Combine π and Fst datasets

The π-ratio and Fst datasets are joined by chromosome, window coordinates, and
comparison so that both statistics are aligned for the same genomic windows.
Only windows present in both datasets are retained.

Rows with missing Fst or log2 π-ratio values are removed, while infinite log2
π ratios are retained for later plotting. A unique identifier is also created
for each window and comparison.

```{r combine-data}
pi_fst_df_merge_clean <- pi_df_ratios %>%
  inner_join(fst_df_prepared,
    by = c("chromosome", "chromo.num", "window_pos_1", "window_pos_2", "comp")) %>%
  filter(!is.na(avg_wc_fst), !is.na(log2_pi.ratio)) %>%
  mutate(chr_pos1_pos2_comp = paste(chromo.num, window_pos_1, window_pos_2, comp, sep = "_")) %>%
  select(chr_pos1_pos2_comp, chromo.num, window_pos_1, window_pos_2, comp, avg_wc_fst, pi_ratio, log2_pi.ratio)

# check to make sure this df looks good so far
# see knitted document version for tutorial example output here
glimpse(pi_fst_df_merge_clean)
```

# 6. Define candidate outlier windows

## Global thresholds

Global thresholds are calculated by pooling windows across all pairwise
comparisons. Before calculating these thresholds, negative Fst estimates are
removed.

Candidate outliers must fall within the upper 1% of Fst values and either the
lower or upper 1% of log2 π ratios. Infinite log2 π ratios are excluded when
estimating the percentile thresholds but are retained when windows are
classified. The resulting table summarizes the number and proportion of global
outliers within each comparison.

These global results are included for exploratory purposes; the meadow plots
use the comparison-specific thresholds calculated in the next section.

```{r global-outliers}
finite_quantile <- function(x, probability) {
  x <- x[is.finite(x)]
  unname(quantile(x, probs = probability, na.rm = TRUE))}

pi_fst_merge_no.neg.fst <- pi_fst_df_merge_clean %>%
  filter(avg_wc_fst >= 0)

global_pi_lower <- finite_quantile(
  pi_fst_merge_no.neg.fst$log2_pi.ratio, 0.01)
global_pi_upper <- finite_quantile(
  pi_fst_merge_no.neg.fst$log2_pi.ratio, 0.99)
global_fst_upper <- unname(quantile(
  pi_fst_merge_no.neg.fst$avg_wc_fst, probs = 0.99, na.rm = TRUE))

pi_fst_merge_no.neg.fst <- pi_fst_merge_no.neg.fst %>%
  mutate(outlier.labels_GLOBAL = if_else(
      avg_wc_fst >= global_fst_upper &
        (log2_pi.ratio <= global_pi_lower | log2_pi.ratio >= global_pi_upper),
      "outlier","not.outlier"))

# summarize global outliers by comparison
global_outlier_summary <- pi_fst_merge_no.neg.fst %>%
  group_by(comp) %>%
  summarise(n_outliers = sum(outlier.labels_GLOBAL == "outlier"), n_total = n(),
            proport_outliers = n_outliers / n_total, .groups = "drop")

global_outlier_table <- global_outlier_summary %>%
  mutate(not.outlier = n_total - n_outliers) %>%
  arrange(match(comp, plot_order)) %>%
  select(comp, n_non.outliers=not.outlier, n_outliers, proport_outliers) %>%
  tibble::column_to_rownames("comp") %>% as.matrix()

global_outlier_table
```

## Comparison-specific thresholds

This section calculates separate thresholds for each pairwise comparison,
allowing the outlier definition to reflect the distribution of values within
that comparison.

Windows are classified as candidate outliers when their Fst is at or above the
comparison-specific 99th percentile and their log2 π ratio is at or below the
1st percentile or at or above the 99th percentile. Infinite log2 π ratios are
excluded when estimating the percentile thresholds but retained during
classification. The final table summarizes the number and proportion of
candidate outlier windows for each comparison.

```{r comparison-outliers}
pi_fst_merge_no.neg.fst_merged_individ.comp <-
  pi_fst_merge_no.neg.fst %>%
  group_by(comp) %>%
  mutate(
    pi_lower.thresh = finite_quantile(log2_pi.ratio, 0.01),
    pi_high.thresh = finite_quantile(log2_pi.ratio, 0.99),
    fst_thresh = unname(quantile(
      avg_wc_fst, probs = 0.99, na.rm = TRUE
    )),
    outlier.labels_individ.comp = if_else(
      avg_wc_fst >= fst_thresh &
        (log2_pi.ratio <= pi_lower.thresh | log2_pi.ratio >= pi_high.thresh),
      "outlier","not.outlier"
    )) %>% ungroup()

# summarize independently defined outliers by comparison
individual_outlier_summary <- pi_fst_merge_no.neg.fst_merged_individ.comp %>%
  group_by(comp) %>%
  summarise(n_outliers = sum(outlier.labels_individ.comp == "outlier"), n_total = n(),
            proport_outliers = n_outliers / n_total, .groups = "drop")

individual_outlier_table <- individual_outlier_summary %>%
  mutate(not.outlier = n_total - n_outliers) %>%
  arrange(match(comp, plot_order)) %>%
  select(comp, n_non.outliers=not.outlier, n_outliers, proport_outliers) %>%
  tibble::column_to_rownames("comp") %>% as.matrix()

individual_outlier_table
```

# 7. Label windows with infinite pi ratios

When π is zero in `population_1`, the log2 π ratio is `-Inf`. When π is zero
in `population_2`, the ratio is `Inf`. These values are biologically informative
and are retained rather than discarded.

This section creates the plotting variables used to distinguish finite and
infinite ratios. Infinite values are replaced with `NA` only in a separate
color-scale column, while the original values remain unchanged. Additional
labels and transparency values allow infinite ratios to be plotted as
asterisks and candidate outliers to remain visually distinct.

```{r label-infinite-values}
main_df_clean <- pi_fst_merge_no.neg.fst_merged_individ.comp %>%
  mutate(
    log2_pi.ratio_inf2NA = if_else(
      is.finite(log2_pi.ratio), log2_pi.ratio, NA_real_
    ),
    tri.coded_inf.or.no = case_when(
      log2_pi.ratio == -Inf ~ "inf_bc.pop1.pi.0", 
      log2_pi.ratio == Inf  ~ "inf_bc.pop2.pi.0",
      TRUE                  ~ "not.inf"
    ),
    outlier.labels_inf.labels = case_when(
      outlier.labels_individ.comp == "outlier" & log2_pi.ratio == -Inf ~ "outlier_pop1.pi.0",
      outlier.labels_individ.comp == "outlier" & log2_pi.ratio == Inf  ~ "outlier_pop2.pi.0",
      outlier.labels_individ.comp == "outlier" ~ "outlier_not.inf",
      log2_pi.ratio == -Inf ~ "not.out_pop1.pi.0",
      log2_pi.ratio == Inf  ~ "not.out_pop2.pi.0",
      TRUE                  ~ "not.out_not.inf"
    ),
    alpha.for.asterisk = case_when(
      outlier.labels_inf.labels %in%
        c("outlier_pop1.pi.0", "outlier_pop2.pi.0") ~ 1,
      outlier.labels_inf.labels %in%
        c("not.out_pop1.pi.0", "not.out_pop2.pi.0") ~ 0.6,
      TRUE ~ 0
    ),
    alpha.for.main.pts = case_when(
      outlier.labels_inf.labels == "outlier_not.inf" ~ 1,
      outlier.labels_inf.labels == "not.out_not.inf" ~ 0.6,
      TRUE ~ 0
    ),
    tri.coded_inf.or.no = factor(
      tri.coded_inf.or.no,
      levels = c("not.inf", "inf_bc.pop1.pi.0", "inf_bc.pop2.pi.0")
    ),
    outlier.labels_individ.comp = factor(
      outlier.labels_individ.comp,
      levels = c("not.outlier", "outlier")
    )
  )
```

## Save the prepared data

This section saves three intermediate datasets that can be inspected or reused
without rerunning the full analysis. The first contains the merged data,
comparison-specific thresholds, and outlier classifications. The second adds
the variables used to create the meadow plots. The third provides a list of
windows with infinite log2 π ratios.

All files are written to the folder specified by `wrangled_dir`.

```{r save-data}
write_csv(pi_fst_merge_no.neg.fst_merged_individ.comp, file.path(wrangled_dir, "pi_fst_outlier_windows_with_infinite_values.csv"))

write_csv(main_df_clean, file.path(wrangled_dir, "pi_fst_outlier_windows_final.csv"))

infinite_window_list <- main_df_clean %>% filter(!is.finite(log2_pi.ratio)) %>% select(chr_pos1_pos2_comp)
write_csv(infinite_window_list, file.path(wrangled_dir, "infinite_pi_ratio_windows.csv"))
```

# 8. Summarize outliers by population

This section assigns each candidate outlier to the population with lower π.
A negative log2 π ratio indicates lower π in `population_1`, while a positive
value indicates lower π in `population_2`. The same rule correctly assigns
negative and positive infinite ratios.

The number of candidate outlier windows assigned to each population is counted
for every comparison and saved to `outlier_counts_by_comparison.csv`.

```{r summarize-outliers}
outlier_counts <- main_df_clean %>%
  filter(outlier.labels_individ.comp == "outlier") %>%
  mutate(
    outlier_population = if_else(
      log2_pi.ratio < 0, "population_1", "population_2"
    )
  ) %>%
  count(comp, outlier_population, name = "n_outlier_windows") %>%
  pivot_wider(
    names_from = outlier_population,
    values_from = n_outlier_windows,
    values_fill = 0,
    names_prefix = "n_outliers_"
  ) %>%
  rename(comparison = comp)

outlier_counts

write_csv(
  outlier_counts,
  file.path(wrangled_dir, "outlier_counts_by_comparison.csv")
)
```

# 9. Create meadow plots

## Color scale

This function creates a separate π-ratio color scale for each comparison. The
scale spans the finite log2 π-ratio values present in that comparison and is
centered at zero, where both populations have equal π. Negative values 
indicate lower π in `population_1`, while positive values indicate lower π 
in `population_2`. 

The hexadecimal color codes can be edited to customize the negative and 
positive sides of the gradient.

Infinite values are excluded from the color scale because they are plotted 
separately as asterisks.

```{r color-scale}
auto_pi_scale <- function(x, title = "Legend", n = 50) {
  x <- x[is.finite(x)]
  limits <- if (length(x) == 0) c(-1, 1) else range(c(x, 0), na.rm = TRUE)
  if (diff(limits) == 0) limits <- limits + c(-1e-9, 1e-9)

  # adjust HEX codes below to modify pi ratio color gradient
  # make sure assigned midpoint value (here, "#92D050" corresp. to pi ratio=0) appears in both (-) & (+) lists
  negative_colors <- c(
    "#3E207B", "#4C3194", "#5A42AD", "#5E63CC", "#5970D2",
    "#5C89DC", "#94BBEC", "#A3CCD0", "#9CCD9E", "#92D050"
  )
  positive_colors <- c(
    "#92D050", "#CAD961", "#FFE171", "#FACA42", "#F3A532",
    "#EB8E2D", "#E2662A", "#BE3D28", "#B12F27", "#752A24"
  )

  zero_position <- scales::rescale(0, from = limits)
  colors <- c(
    colorRampPalette(negative_colors)(ceiling(n / 2)),
    colorRampPalette(positive_colors)(ceiling(n / 2))[-1]
  )
  values <- c(
    seq(0, zero_position, length.out = ceiling(n / 2)),
    seq(zero_position, 1, length.out = ceiling(n / 2))[-1]
  )

  scale_color_gradientn(
    name = title,colors = colors, values = values,
    limits = limits, oob = scales::squish)
}
```

## Plot function

This function builds one meadow plot for each comparison. Fst is plotted across
chromosomes, with finite log2 π ratios shown using the comparison-specific color
scale and infinite values shown as colored asterisks. Candidate outliers are
displayed as larger, fully opaque points.

Comments within the function indicate where plot colors, point sizes, text, and 
legend settings can be customized.

```{r plot-function}
make_meadow_components <- function(comparison_name) {
  df <- main_df_clean %>%
    filter(comp == comparison_name)

  y_max <- max(df$avg_wc_fst, na.rm = TRUE)
  y_limit <- if (y_max > 0) y_max * 1.05 else 1
  
  pop_names <- strsplit(comparison_name, ".v.", fixed = TRUE)[[1]]
  legend_title <- paste0(
    "\u03c0 ratio<br>",
    "<span style='color:#E44E20; font-weight:bold'>", # adjust pop 2 legend title color here
        pop_names[2], "</span><br>",
    "vs<br>",
    "<span style='color:#545FBD; font-weight:bold'>", # adjust pop 1 legend title color here
        pop_names[1], "</span>"
  )
  
  common_facet <- facet_grid(~ chromo.num, scales = "free_x", space = "free_x", switch = "x")

  common_theme <- theme(
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 8, color = "black"), # edit appearance of legend numeric values here 
    axis.ticks.x = element_blank(),
    axis.title = element_blank(),
    panel.spacing.x = grid::unit(0.03, "lines"),
    plot.margin = margin(t = 2.75, r = 5.5, b = 2.75, l = 5.5, unit = "pt"))

  main_points <- ggplot(df, aes(window_pos_1, avg_wc_fst)) +
  geom_point(data = filter(df, is.finite(log2_pi.ratio)),
             aes(color = log2_pi.ratio, size = outlier.labels_individ.comp,
                 alpha = alpha.for.main.pts)) +
  geom_point(data = filter(df, log2_pi.ratio == -Inf),
             aes(size = outlier.labels_individ.comp, alpha = alpha.for.asterisk),
             shape = 8, color = "#6579D7") + # adjust -inf asterisk data point color here
  geom_point(data = filter(df, log2_pi.ratio == Inf),
             aes(size = outlier.labels_individ.comp, alpha = alpha.for.asterisk),
             shape = 8, color = "#D55C2C") + # adjust +inf asterisk data point color here
  auto_pi_scale(df$log2_pi.ratio, title = legend_title) +
  scale_alpha_identity() +
  scale_size_manual(values = c("not.outlier" = 0.3, "outlier" = 1.6),guide = "none") + # adjust data point size here
  scale_y_continuous(limits = c(0, y_limit), expand = expansion(mult = c(0, 0.01))) +
  common_facet + transparent_theme + common_theme +
  theme(legend.position = "right",
        legend.title = ggtext::element_markdown(hjust = 0, size = 7, lineheight = 1),
        legend.text = element_text(size = 7, color = "black", # edit appearance of legend numeric values here 
                                   margin = margin(l = 3, unit="pt")), 
        legend.background = element_rect(fill = "transparent", color = NA),
        legend.box.background = element_rect(fill = "transparent", color = NA),
        legend.key = element_rect(fill = "transparent", color = NA),
        legend.key.height = grid::unit(0.8, "cm"),legend.key.width = grid::unit(0.3, "cm")#,
        #legend.ticks = element_blank()
        ) + 
  guides(color = guide_colorbar(title.position = "top", barwidth = grid::unit(0.4, "cm"),
                                frame.colour = "black", frame.linewidth = 0.3,
                                ticks.colour = "black",ticks.linewidth = 0.15))
  list(main_points = main_points)
}

meadow_plots <- setNames(lapply(comps, make_meadow_components), comps)
```

The completed plots are stored in `meadow_plots` and can be accessed by
comparison name. To preview an individual plot, uncomment the line below and
replace `"Sal.v.SPW"` with a value from the `comp` column of your comparison
map.

```{r example-plot, fig.width=12, fig.height=4}
# meadow_plots[["Sal.v.SPW"]]$main_points
```

# 10. Combine and save the plots

The individual meadow plots are arranged vertically according to `plot_order`.
Chromosome labels and the x-axis title are shown only on the bottom plot, while
each comparison retains its own y-axis scale and π-ratio legend. 

The final figure is saved at 600 dpi. Its height increases with the number of
comparisons but is capped at 60 cm.

```{r combine-plots}
remove_chromosome_strips <- function(plot_object) {
  plot_object + theme(strip.background = element_blank(),strip.text = element_blank())}

add_combined_panel_border <- function(plot) {
  grob <- ggplotGrob(plot)
  panels <- grob$layout[grepl("^panel", grob$layout$name), ]

  gtable::gtable_add_grob(
    grob,
    grid::rectGrob(gp = grid::gpar(
      color = "black", lwd = 1)), # edit panel border color here
    t = min(panels$t), b = max(panels$b),
    l = min(panels$l), r = max(panels$r),
    z = Inf, clip = "off"
  )
}

main_point_panels <- lapply(seq_along(plot_order), function(i) {
  plot <- meadow_plots[[plot_order[i]]]$main_points + labs(x = "Chromosome")

  if (i < length(plot_order)) {
    plot <- plot + theme(strip.background = element_blank(),
                         strip.text.x = element_text(color = "transparent", size = 1),
                         axis.title.x = element_text(color = "transparent", size = 1))
  } else {
    plot <- plot + theme(strip.background = element_blank(),
                         strip.text.x = element_text(),
                         axis.title.x = element_text(size = 9, color = "black")) # edit shared x-axis label here
  }

  add_combined_panel_border(plot)
})

shared_y_title <- grid::textGrob(
  expression(paste(
    "Average ", italic(F)[italic(ST)], "(15-kb windows, 5-kb step)")), # edit shared y-axis label text here
  rot = 90,
  gp = grid::gpar(fontsize = 9) # edit shared y-axis label text size here
)

meadow.plots_main.pts_combined <- arrangeGrob(
  grobs = main_point_panels, ncol = 1,
  left = shared_y_title)
```

The combined figure is saved as a transparent PNG in the folder specified by
`plot_dir`. 

```{r save-plots}
ggsave(
  file.path(plot_dir, "meadow_plots.png"),
  meadow.plots_main.pts_combined,
  width = 20, # edit final figure width here
  height = min(8*nrow(comparison_map),60), # edit final figure height here
  units = "cm", dpi = 600,
  bg = "transparent"
)
```


# 11. Group candidate outlier windows into peaks

The meadow plots identify individual candidate outlier windows, but a broader
candidate region may contain several nearby outlier windows. This section groups
those windows into candidate peaks and counts the number of outlier windows in
each peak.

The original analysis allowed neighboring outlier windows to remain in the same
peak when they were separated by no more than three retained non-outlier windows.
That value can be changed below. Windows removed earlier because of missing data
or negative Fst are not counted as intervening non-outlier windows.

```{r candidate-peak-settings}
max_intervening_nonoutliers <- 3

if (length(max_intervening_nonoutliers) != 1 ||
    is.na(max_intervening_nonoutliers) ||
    max_intervening_nonoutliers < 0 ||
    max_intervening_nonoutliers %% 1 != 0) {
  stop("max_intervening_nonoutliers must be one non-negative whole number.")
}
```

## Index retained windows

Each retained window is assigned a sequential index within its chromosome and
comparison. Using this index reproduces the original binary-sequence approach
without converting the outlier labels to character strings or counting peaks
manually.

```{r index-windows}
indexed_windows <- main_df_clean %>%
  arrange(comp, chromo.num, window_pos_1, window_pos_2) %>%
  group_by(comp, chromo.num) %>%
  mutate(retained_window_index = row_number()) %>%
  ungroup()
```

## Define candidate peaks

After non-outlier windows are removed from the indexed table, consecutive
outliers are assigned to the same peak unless their retained-window indices
differ by more than the allowed gap. The resulting table contains one row per
candidate peak, including its coordinates, number of outlier windows, number of
intervening retained non-outlier windows, and total genomic span.

```{r define-candidate-peaks}
candidate_peaks <- indexed_windows %>%
  filter(outlier.labels_individ.comp == "outlier") %>%
  group_by(comp, chromo.num) %>%
  arrange(retained_window_index, .by_group = TRUE) %>%
  mutate(
    new_peak = is.na(lag(retained_window_index)) |
      retained_window_index - lag(retained_window_index) >
        max_intervening_nonoutliers + 1,
    peak_number = cumsum(new_peak)
  ) %>%
  group_by(comp, chromo.num, peak_number) %>%
  summarise(
    first_window_start = min(window_pos_1),
    last_window_end = max(window_pos_2),
    n_outlier_windows = n(),
    n_intervening_retained_nonoutliers =
      max(retained_window_index) - min(retained_window_index) + 1 - n(),
    total_genomic_span_kb =
      (max(window_pos_2) - min(window_pos_1) + 1) / 1000,
    .groups = "drop"
  ) %>%
  arrange(match(comp, plot_order), chromo.num, first_window_start)

if (nrow(candidate_peaks) == 0) {
  stop("No candidate outlier windows were available for peak analysis.")
}

glimpse(candidate_peaks)
```

# 12. Summarize and save candidate peaks

The comparison summary reports the total number of candidate peaks and outlier
windows, along with the mean and maximum number of outlier windows per peak.
The peak-size table reports how many peaks of each size occur within each
comparison. All three candidate-peak tables are saved in `wrangled_dir`.

```{r summarize-candidate-peaks}
comparison_peak_summary <- candidate_peaks %>%
  group_by(comp) %>%
  summarise(
    n_candidate_peaks = n(),
    mean_outlier_windows_per_peak = mean(n_outlier_windows),
    max_outlier_windows_per_peak = max(n_outlier_windows),
    n_outlier_windows = sum(n_outlier_windows),
    .groups = "drop"
  ) %>%
  relocate(n_outlier_windows, .after = n_candidate_peaks) %>%
  arrange(match(comp, plot_order))

peak_size_summary <- candidate_peaks %>%
  count(comp, n_outlier_windows, name = "n_candidate_peaks") %>%
  arrange(match(comp, plot_order), n_outlier_windows)

comparison_peak_summary
peak_size_summary

write_csv(candidate_peaks, file.path(wrangled_dir, "candidate_peaks.csv"))
write_csv(comparison_peak_summary, file.path(wrangled_dir, "candidate_peak_summary_by_comparison.csv"))
write_csv(peak_size_summary, file.path(wrangled_dir, "candidate_peak_size_counts.csv"))
```

# 13. Plot candidate peak sizes

Each point represents one candidate peak, and its vertical position gives the
number of outlier windows assigned to that peak. The violin layer summarizes
the distribution of peak sizes within each comparison. Comparisons follow the
same `plot_order` used for the meadow plots.

```{r plot-candidate-peaks}
candidate_peaks_for_plot <- candidate_peaks %>%
  mutate(comp = factor(comp, levels = plot_order))

candidate_peak_size_plot <- ggplot(
  candidate_peaks_for_plot,
  aes(comp, n_outlier_windows, group = comp)) +
  
  geom_violin(width = 1, 
              color = "#797979", # violin plot outline color
              fill = "#797979",  # violin plot fill color
              alpha = 0.25) +  # violin plot fill transparency
  
  geom_jitter(aes(size = n_outlier_windows), 
              color = "black", # individual point outline color
              fill = "#797979",  # individual point fill color
              alpha = 0.9, # individual point fill transparency
              width = 0.25, # width of point jitter zone
              shape = 21) + # shape of point (21 = filled circle)
 
  scale_size(range = c(1.5, 3), # scale point size to reduce crowding at low values
             guide = "none") +
  scale_x_discrete(drop = FALSE, labels = function(x) gsub(".v.", "\nvs\n", x, fixed = TRUE)) +
  scale_y_continuous(breaks = scales::breaks_pretty(n = 4)) +
  labs(x = "Population comparison", y = "Outlier windows per candidate peak") +
  theme_classic() +
  transparent_theme +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, lineheight = 0.9),
      panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5))

candidate_peak_size_plot
```

The candidate-peak plot is saved as a transparent PNG in `plot_dir`. Its
filename, dimensions, resolution, and background can be adjusted below.

```{r save-candidate-peak-plot}
ggsave(
  file.path(plot_dir, "candidate_peak_size_distributions.png"),
  candidate_peak_size_plot,
  width = min(5*nrow(comparison_map),60), # edit final figure width here
  height = 15, # edit final figure height here
  units = "cm", dpi = 600,
  bg = "transparent"
)
```

# 14. Reproducibility information

`sessionInfo()` records the R version, operating system, and package versions
used to run the tutorial. See the knitted document version for the session
information under which this tutorial was built.

```{r session-info}
sessionInfo()
```
